Experience Report: Haskell in Computational Biology

### 2012-02-28 mrfy_experience_report.tex
Break out viterbi types.
Add some notes on connecting the types to the concepts.
Another figure and further explanations. We'll need to shrink some code.

SMURF, SMURFLite, and MRFy all attempt to compute the optimal parse, or 
alignment, of a protein sequence onto the directed graph representing a Markov 
random field (see Figure \ref{mrf}). The algorithm underlying MRFy grew out of 
an understanding of the limitations of the SMURF (Structural Motifs Using 
Random Fields) algorithm, which used multidimensional dynamic programming to 
compute the optimal parse of a query sequence onto a Markov random field. 
Rather than attempt to compute exactly the optimal parse, MRFy treats the 
non-local interactions (which correspond to hydrogen-bonded ``beta strands'' in 
the protein) as ``beads'' which can be slid along a string. In this analogy, 
the string is the sequence of amino acids of the query protein. A given 
placement of these beads corresponds to an assignment of particular amino acid 
residues being hydrogen bonded to other particular amino acid residues, and 
thus MRFy computes a probability based on observed frequencies of residues 
hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
strands, the Markov random field is in fact a traditional hidden Markov model, 
whose optimal solution can be easily computed via dynamic programming (known as 
the Viterbi algorithm). MRFy performs a stochastic search to place these beta 
strands, based on an objective function that combines the probability computed 
by the Viterbi algorithm with that computed from the paired, hydrogen-bonded 
residues in the beta strands.

@@

\subsection{Beautiful code}
MRFy contains two distinct algorithms: the classical Viterbi dynamic 
programming algorithm for hidden Markov models, and a stochastic search 
algorithm for placing beta strand positions. This stochastic search algorithm 
implements several search heuristics: random-mutation hill climbing, simulated 
annealing, multistart simulated annealing, and a genetic algorithm.

The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
probability of an observed sequence (in this case, of amino acid ``letters'') 
being emitted by a particular finite-state automaton (the hidden Markov model). 
The particular states of this finite-state automaton map to the recurrence 
relations, and in this instance derive from what is known in the computational 
biology community as a \textit{profile} hidden Markov model. The Viterbi 
algorithm in this instance recurses on ``match,'' ``insertion,'' and 
``deletion'' states. To be efficient, the Viterbi algorithm relies on dynamic 
programming, or memoization, and the resulting asymptotic complexity is 
$O(|M|\times|N|)$, when $M$ is the sequence of distinct states (nodes in the 
automaton's graph) and $N$ is the sequence of letters in the protein sequence. 
In our particular problem domain, both $M$ and $N$ routinely contain several 
hundred to a few thousand elements.

@@

We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
bottom-up dynamic programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model. The array containing the paths 
served as the memoization table. We next wrote a top-down dynamic programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both). In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array). We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{durbin} or the HMMER literature~\cite{eddy} 
(see Equation \ref{viterbi_eqn}). We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases. This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node. In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down dynamic 
programming approach in Haskell, which resulted in perspicuous code.

@@

@@

>>>>

SMURF, SMURFLite, and MRFy all attempt to compute the optimal parse, or 
alignment, of a protein sequence onto the directed graph representing a Markov 
random field (see Figure \ref{mrf}). A parse is a mapping of the letters of a 
query sequence onto the states of a markov Random field, while an optimal parse 
is the parse exhibiting the greatest probability once. The algorithm underlying 
MRFy grew out of an understanding of the limitations of the SMURF (Structural 
Motifs Using Random Fields) algorithm, which used multidimensional dynamic 
programming to compute the optimal parse of a query sequence onto a Markov 
random field. Rather than attempt to compute exactly the optimal parse, MRFy 
treats the non-local interactions (which correspond to hydrogen-bonded ``beta 
strands'' in the protein) as ``beads'' which can be slid along a string. In 
this analogy, the string is the sequence of amino acids of the query protein. A 
given placement of these beads corresponds to an assignment of particular amino 
acid residues being hydrogen bonded to other particular amino acid residues, 
and thus MRFy computes a probability based on observed frequencies of residues 
hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
strands, the Markov random field is in fact a traditional hidden Markov model, 
whose optimal solution can be easily computed via dynamic programming (known as 
the Viterbi algorithm). MRFy performs a stochastic search to place these beta 
strands, based on an objective function that combines the probability computed 
by the Viterbi algorithm with that computed from the paired, hydrogen-bonded 
residues in the beta strands.

@@

\subsection{Beautiful code}
MRFy contains two distinct algorithms: the classical Viterbi 
dynamic-programming algorithm for hidden Markov models, and a stochastic search 
algorithm for placing beta-strand positions. This stochastic search algorithm 
implements several search heuristics: random-mutation hill climbing, simulated 
annealing, multistart simulated annealing, and a genetic algorithm.

The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
probability of an observed sequence (in this case, of amino acid ``letters'') 
being emitted by a particular finite-state automaton (the hidden Markov model). 
The particular states of this finite-state automaton map to the recurrence 
relations, and in this instance derive from what is known in the computational 
biology community as a \textit{profile} hidden Markov model. The Viterbi 
algorithm in this instance recurses on ``insertion,'' ``deletion,'' and 
``match'' states (Figure~\ref{plan7}). Insertion states represent mutations 
that gain an amino acid letter in the query sequence, deletion states represent 
mutations that lose an amino acid letter in the query sequence, and match 
states represent either conserved positions (no change) or mutations that 
substitute one amino acid for another. The match and insertion states each have 
an associated table of emission probabilities, determined during training, 
indicating how likely that state is to emit a particular amino acid letter. 
Edges from state to state represent state transitions, and have associated 
probabilities determined during training. To be efficient, the Viterbi 
algorithm relies on dynamic programming, or memoization, and the resulting 
asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence. In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements. The types of 
these states and probabilities are provided in Figure~\ref{viterbi_types}.

\begin{figure*}
\lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
\end{figure*}

\begin{figure} \centerline{\includegraphics[width=8cm]{Plan7.pdf}} \caption{The 
directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
Markov model. Only 7 of the possible 9 state transitions are allowed, which 
simplifies the dynamic programming. Insertion states represent mutations that 
insert an amino acid, deletion states represent mutations that lose an amino 
acid, and match states represent either conserved amino acids (those with no 
changes) or mutations that substitute one amino acid for another. Each match or 
insertion state has an associated table of emission probabilities (here 
illustrated only for the first two states)}\label{plan7} \end{figure}

The textbook Viterbi recurrence relations refer to logs of probabilities, so 
that the terms may be added rather than multiplying raw probabilities. For 
reasons including floating point underflow, the HMMER software (with which we 
maintain file format compatibility) stores all probabilities in a trained HMM 
file as negative natural logs. Thus, the Viterbi recurrence relations are 
simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
and because they are \textit{negative} logs, the problem transforms from 
maximization to minimization.

@@

@@

We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model. The array containing the paths 
served as the memoization table. We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both). In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array). We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{durbin} or the HMMER literature~\cite{eddy} 
(see Equation \ref{viterbi_eqn}). We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases. This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node. In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down 
dynamic-programming approach in Haskell, which resulted in perspicuous code.

@@

We note that \texttt{seed} in \texttt{search} comes from a monadic random 
number generator in Main, but we wished for search itself to be pure at this 
level. One advantage is that this allows for testing of the \texttt{search} 
function with no randomness at all: if we pass in a known number as the seed, 
behavior is entirely deterministic.
